Changes in accessibility to food banks during COVID-19 and implications for the food security of vulnerable populations in Hamilton, Ontario

### 2021-04-23 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.tex
Added LISA analysis of changes in accessibility and rebuilt package to include Hamilton boundary that I had mistakenly deleted before

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Changes in accessibility to foodbanks during COVID-19 and implications for the food security of vulnerable populations in Hamilton, Ontario},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Changes in accessibility to foodbanks during COVID-19 and
implications for the food security of vulnerable populations in
Hamilton, Ontario}
    \author[Some University]{Author One\corref{1}}
   \ead{author1@example.com} 
    \author[]{Author Two}
   \ead{author2@example.com} 
    \author[Some University]{Author Three\corref{2}}
   \ead{author3@example.com} 
    \author[Another University]{Author Four\corref{2}}
   \ead{author4@example.com} 
      \address[Some University]{Department, Street, City, State, Zip}
    \address[Another University]{Department, Street, City, State, Zip}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

@@

Food insecurity is defined as an ``inadequate or uncertain access to a
sufficient quantity and/or adequate quality of food'' due to a
household's financial limitations (Enns et al., 2020). This condition
has been associated with reductions in nutritional outcomes
(Bhattacharya et al., 2004; Kirkpatrick and Tarasuk, 2008; Olson, 1999)
and physical and mental health in children and adults (Elgar et al.,
2021; Jones, 2017; Ramsey et al., 2011; Seligman et al., 2010; Stuff et
al., 2004). Over at least the past four decades foodbanks have become an
essential line of defense against food insecurity in Canadian
communities (Black and Seto, 2020; Holmes et al., 2018; Riches, 2002;
Tarasuk et al., 2020). In this respect, Canada is not unlike numerous
other wealthy countries where a systematic dismantling of the welfare
state took place in the intervening period (Tarasuk et al., 2014).

@@

The impacts of food insecurity during the pandemic are alarming, since
diet-related diseases, such as obesity, heart-disease, and diabetes,
were already critical public health concerns in Canada prior to COVID-19
(Boucher et al., 2017). While foodbanks are not necessarily a stable
solution to food insecurity and in fact may encourage a retrenchment of
neoliberal policy (Wakefield et al., 2013), at least they can be argued
to provide a resource of last instance to households in precarious
situations (Bazerghi et al., 2016). As recently as 2019, the Hamilton
Hunger Report\footnote{https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf}
noted that foodbanks in Hamilton, Ontario, recorded the highest number
of visitors in the past 29 years; the number of children visiting
foodbanks (minors up to 18 years old) was 9,125 in March 2019, up from
8,278 the year before, a rate of increase greater than population
growth.

It is known that urban food environments, within which people make their
daily food choices, are essential in influencing eating behaviours and
health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to foodbanks. Although
previous work has explored differences in \emph{accessing} foodbanks,
such as how some households utilize foodbanks over short periods of time
while others regularly utilize foodbanks as longer-term resource (e.g.,
Enns et al., 2020), we are not aware of any research that has focused on
the geographic component of accessibility.

The study of place-based geographic accessibility is concerned with
capturing the potential to reach destinations of value using the
transportation network (Páez et al., 2012). Indeed, the Government of
Canada's recent Food Policy has made ``access'' to healthy food a
priority for Canadian communities\footnote{https://www.agr.gc.ca/eng/about-our-department/key-departmental-initiatives/food-policy/the-food-policy-for-canada/?id=1597863791042}
and previous survey research has suggested that such accessibility plays
a key role in user satisfaction with foodbank service delivery (Holmes
et al., 2018). However, as with research into the prevalence of food
deserts, accessibility to foodbanks is unlikely to be evenly
distributed, and variation throughout a city can be expected due to
transportation network characteristics, and the spatial distribution of
foodbank locations and the population they are meant to serve.
Furthermore, policy responses to the COVID-19 pandemic likely have added
to the distress of vulnerable households. Non-pharmaceutical
interventions during the pandemic involving restrictions in mobility
have increased the friction of travel, in particular by transit on which
low income populations are more reliant (e.g., DeWeese et al., 2020). At
the same time, the pandemic has created additional stress for the
operators of foodbanks through disruptions in the supply chain (e.g.,
McKay et al., 2021) as well as concerns surrounding the delivery of
service in safe conditions and possible cancellation of food service
programs.

For this study, we aim to look at how the landscape of foodbanks and
related services (e.g.~low-cost or free meal service providers)
available in Hamilton, Ontario, changed during the pandemic. Did the
number of open foodbank services diminish? If so, what was the
accessibility to foodbanks before the pandemic from the perspective of
low income households, and how it changed during the pandemic? And
finally, who are most likely to have been impacted by changes in the
accessibility landscape? This paper first looks at the distribution of
foodbanks and related services before and during the pandemic. Then, we
use the balanced floating catchment area approach of Paez et al. (2019)
to investigate the accessibility situation. For this, we use a fully
disaggregated approach based on parcel-level data. Socio-economic and
demographic data are drawn from the latest Census of Canada (2016),
whereas travel information is from the most recent regional travel
survey from 2016. This paper follows reproducible research
recommendations (see Brunsdon and Comber, 2020), and the research was
conducted using open source tools for transportation analysis (Lovelace,
2021). The code and data necessary to reproduce the analysis are
available in a public repository\footnote{add repository}.

\hypertarget{food-insecurity-and-foodbank-use-in-canada}{%
\section{Food Insecurity and foodbank Use in
Canada}\label{food-insecurity-and-foodbank-use-in-canada}}

@@

Foodbanks - sometimes also referred to as `food pantries' and `food
shelves' - originated as a community response to aid those with
inadequate food by voluntarily offering them meals and ingredients
(Loopstra and Tarasuk, 2012; Riches, 2002). The scope and objectives of
foodbanks can vary by region and by country, and these organizations can
include not only prepared meals and aliments, but also shared spaces to
connect in community gardens and community kitchens (Wakefield et al.,
2013). Although in their origin foodbanks were meant to be provide a
temporary solution to accommodate those in hunger due to job
retrenchments and economic downfalls since the 1980s, over time they
have evolved into a community practice to secure emergency food supplies
for those in need (Loopstra and Tarasuk, 2012; Wakefield et al., 2013).

In Canada, the number of foodbanks has steadily increased in the past
few decades (Wakefield et al., 2013). The largest database of foodbanks
and their use comes from the non-profit association Foodbanks Canada,
which conducts an annual assessment through its affiliated members.
Foodbanks Canada's 2018 Hunger Count report\footnote{https://foodbankscanada.ca/getmedia/241fb659-05f5-44a2-9cef-56f5f51db523/HungerCount-2018\_FINAL\_EN.pdf.aspx?ext=.pdf}
(the most recent available) listed 1,830 member foodbanks across the
country, and found that Canadians visited foodbanks 1.1 million times in
March of 2018. Of those accessing foodbanks, certain population
characteristics tend to be over-represented compared to national totals
from the 2016 Canadian Census of Population. According to Foodbanks
Canada's 2018 data, single-adult households represent 45\% of those
utilizing foodbanks despite making up 28\% of Canada's population, 19\%
are single-parent households (compared to 10\% nationally), and 35\% of
those accessing foodbank services are children aged 0-18 even though
their share of Canada's national population is approximately 20\%. In
addition, 59\% of households accessing foodbanks list social or
disability assistance as their primary source of income. Moreover,
children aged 0-18 are over-represented as foodbank users, making up
35\% of those accessing foodbank services despite representing 20\% of
the country's population. Similarly, using data from the 2011-2012 CCHS,
Tarasuk et al. (2019) found higher odds of food insecurity amongst
households relying on social assistance, those without a university
degree or with children under the age of 18, and individuals that lived
alone, renters, and those identifying as Aboriginal. While surveys
revealed that only 20 to 30 percent of those experiencing food
insecurity were found to frequent foodbanks in Canada (Tarasuk et al.,
2014), research from Ottawa (Enns et al., 2020) and Vancouver (Black and
Seto, 2020) suggests that long-term users tend to be older, have health
or mobility challenges, live in large households, and are less likely to
have employment income.

Economic stress caused by the COVID-19 pandemic, rising unemployment
rates, and changes in poverty levels have disrupted the food environment
and led to higher rates of food insecurity (Niles et al., 2020).
Pandemic-related food security studies in the US have found a
substantial increase in households experiencing food insecurity for the
first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects
associated with food insecurity, increases in food insecurity rates due
to the pandemic, signal a change in the food environment with potential
damages to population health during the course of the pandemic and
beyond (Niles et al., 2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from Foodbanks Canada showed that 52\% of member foodbanks reported an
increase in usage in March of 2020 when initial lockdown restrictions
were put in place across much of the country. The pandemic also created
significant staffing issues with 42\% of foodbanks reporting a reduction
in volunteers. However, 53\% of foodbanks reported a decrease in use
into the summer of 2020 which foodbanks attributed to emergency
financial support programs from the federal government were.
Nevertheless, some of these benefit programs were temporary, which
suggests that many households may again turn to foodbank services to
meet their needs.

In terms of geography, previous research conducted at the provincial
scale using data from the 2011-2012 CCHS found that the prevalence of
food insecurity ranged across the country from 11.8\% of households in
Ontario to 41\% of households in Nunavut (Tarasuk et al., 2019).
However, although previous research has examined the characteristics of
individuals and households accessing foodbanks, the locational or
transportation accessibility aspect of foodbank access is not well
understood. A wealth of literature examining the food desert concept
suggests that, in addition to socio-economic and demographic factors,
location and transportation networks play a key role in a household's
accessibility to healthy foods (Paez et al., 2010; Vanderlee and L'Abbé,
2017; Widener, 2018). For foodbanks specifically, previous qualitative
research in Ontario has noted difficulties associated with
transportation for trips to the foodbank (Smith-Carrier et al., 2017).
For example, Smith et al. (2017) note that participants access foodbanks
using a variety of modes of transportation with respondent stating ``I
wish it {[}the foodbank{]} was a little more centrally located. Because
if I didn't have a bike I'd have to walk it all the way out there and
back. I wonder about people who don't.'' From this, the authors conclude
that ``transportation can be challenging, particularly if the foodbank
is situated in a remote location.'' To offer greater insight into the
role of transportation and location in foodbank accessibility, this
research examines how geographic accessibility to foodbanks and food
services changed in Hamilton during the COVID-19 pandemic.

@@

\hypertarget{destinations-foodbanks-and-food-service-locations}{%
\subsubsection{Destinations: foodbanks and Food Service
Locations}\label{destinations-foodbanks-and-food-service-locations}}

The locations of foodbanks and related food services were obtained from
the Hamilton Public Library's Food Access Guide\footnote{http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf}.
The guide was updated in April of 2021 to indicate any change affected
on the services due to the pandemic. This includes modified business
hours, a need to make reservations before frequenting, and locations
that have completely shut down in consequence. While some foodbank
services have a specific a target population, such as prioritizing
family with young children aged between 0 and 3 or accepting only those
providing proof of low-income status through housing and utility
statements, all the foodbank services indicated below are designed to
accommodate those in need of food at zero to low cost. With our focus on
foodbanks and food services that offer free or low-cost meals at
particular locations, we first removed services such as Meals on Wheels
and other food access services such as food box, community kitchens,
student nutrition programs, and shopping and transportation. In
addition, two free meal services held on different days at the same
location were collapsed into a single service point for the
accessibility analysis.

@@

Travel time tables for three modes (car, transit, walking) were computed
using the parcels as the origins and the locations of the foodbanks as
the destinations. For routing, the package \texttt{r5r} (Pereira et al.,
2021b) was used with a network extract for the City of Hamilton from
OpenStreetMaps and the General Transit Feed Specification from Hamilton
Street Railway, the local transit operator. For routing purposes we used
maximum values of 180 min and 10,000 m walking distance: any destination
that exceeded these thresholds was ignored. The departure time used for
routing was \textbf{TIME}.

@@

Once we obtained travel time tables with population (number of
households) and proportion of trips by mode, we calculated the expected
travel time \(ett\) from each parcel \(i\) to a foodbank or food service
location \(j\) as follows: \[
ett_{ij} = p^c_i\cdot tt^c_{ij} + p^t_i\cdot tt^t_{ij} + p^w_i\cdot tt^w_{ij}
\]

\noindent where \(p^k_i\) is the proportion of trips by mode \(k\) in
the TAZ of parcel \(i\), and \(tt^k_{ij}\) is the travel time from
parcel \(i\) to the foodbank. In other words, the expected travel time
is the weighted sum of travel times to the foodbank, with the weights
given by the expected modal split in the TAZ.

@@

Figure \ref{fig:foodbanks} shows the location of foodbanks in the City
of Hamilton and their status. Before the pandemic there were 58 of which
14 (24.14\%) closed during the pandemic. As seen in the figure,
foodbanks tend to be predominantly located in the central parts of the
city. This is not surprising: population density is high there, and it
is also the part of the city where lower income households are more
numerous in absolute and relative terms (see Figure
\ref{fig:low-income-households}). Alas, this is also the part of the
city where most of the closures during the pandemic happened.

@@

To implement the accessibility calculations, we must select a
distance-decay function. In this task we find limited support in the
literature, which is mostly silent on the travel patterns of people who
visit foodbanks. For this reason, we opt for a simple cumulative
opportunities function as follows: \[
w_{ij}=w_{ji}=
\begin{cases}
1 & \text{ if } ett_{ij}\le \delta\\
0 & \text{ otherwise}
\end{cases}
\] \noindent where \(ett_{ij}\) is the multimodal expected travel time
as described previously, and \(\delta\) is a travel threshold. When the
expected travel time exceeds this threshold, a facility is no longer
considered accessible. Further, the weights are standardized for the
balanced floating catchment area approach.

Figure \ref{fig:sensitivity-analysis} shows the results of conducting
sensitivity analysis of the system-wide accessibility as we vary the
threshold (considering the situation before the pandemic). There is a
clear pattern whereby more strict values of \(\delta\) are associated
with higher levels of system-wide accessibility: this is a result of
lower congestion, since fewer households are serviced. System-wide
accessibility declines with higher values of \(\delta\): as more
households are serviced, congestion grows and the level of service
declines, although this happens at a declining rate. We are not aware of
any standards that explain how long people are expected to travel for
foodbanks, but we note that in developing countries, access to drinking
water is defined as a source that takes less than 30 minutes to reach
(round trip, see UNICEF-WHO, 2019). There is no reason why people in
affluent countries should be expected to travel more for a basic need
such as food. Accordingly, we adopt a threshold of 15 minutes for the
analysis (representing a one-way trip). Using this threshold, we find
that the system-wide accessibility (interpreted as a
provider-to-population ratio) was 0.065 (foodbanks per low income
household in the city) before COVID-19, but declined to 0.037 during the
pandemic. It is striking that although almost 76\% of foodbanks remained
in operation during the pandemic, there was a loss of accessibility
greater than 43\%.

@@

The level of service by foodbank before and during the pandemic are
shown in Figure \ref{fig:levels-of-service}. Since the level of service
is the inverse of the number of households in the catchment area of the
foodbanks, higher values mean that a facility is expected to service
fewer households. Conversely, lower values indicate grater congestion.
The general pattern of the levels of service is similar before and
during the pandemic, with lower values in the center of the city. Three
more peripheral facilities towards the south of the city also have lower
levels of service, presumably because they are meant to service a
relatively large suburban/exurban population. During the pandemic,
however, the levels of service dropped, in some cases quite
substantially. The pattern of the losses in level of service, moreover,
is not uniform. As seen in Figure \ref{fig:levels-of-service-changes}
\ldots{} words.

@@

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Changes in accessibility to food banks during COVID-19 and implications for the food security of vulnerable populations in Hamilton, Ontario},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Changes in accessibility to food banks during COVID-19 and
implications for the food security of vulnerable populations in
Hamilton, Ontario}
    \author[Some University]{Author One\corref{1}}
   \ead{author1@example.com} 
    \author[]{Author Two}
   \ead{author2@example.com} 
    \author[Some University]{Author Three\corref{2}}
   \ead{author3@example.com} 
    \author[Another University]{Author Four\corref{2}}
   \ead{author4@example.com} 
      \address[Some University]{Department, Street, City, State, Zip}
    \address[Another University]{Department, Street, City, State, Zip}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

@@

Food insecurity is defined as an ``inadequate or uncertain access to a
sufficient quantity and/or adequate quality of food'' due to a
household's financial limitations (Enns et al., 2020). This condition
has been associated with reductions in nutritional outcomes
(Bhattacharya et al., 2004; Kirkpatrick and Tarasuk, 2008; Olson, 1999)
and physical and mental health in children and adults (Elgar et al.,
2021; Jones, 2017; Ramsey et al., 2011; Seligman et al., 2010; Stuff et
al., 2004). Over at least the past four decades food banks have become
an essential line of defense against food insecurity in Canadian
communities (Black and Seto, 2020; Holmes et al., 2018; Riches, 2002;
Tarasuk et al., 2020). In this respect, Canada is not unlike numerous
other wealthy countries where a systematic dismantling of the welfare
state took place in the intervening period (Tarasuk et al., 2014).

@@

The impacts of food insecurity during the pandemic are alarming, since
diet-related diseases, such as obesity, heart-disease, and diabetes,
were already critical public health concerns in Canada prior to COVID-19
(Boucher et al., 2017). While food banks are not necessarily a stable
solution to food insecurity and in fact may encourage a retrenchment of
neoliberal policy (Wakefield et al., 2013), at least they can be argued
to provide a resource of last instance to households in precarious
situations (Bazerghi et al., 2016). As recently as 2019, the Hamilton
Hunger Report\footnote{https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf}
noted that food banks in Hamilton, Ontario, recorded the highest number
of visitors in the past 29 years; the number of children visiting food
banks (minors up to 18 years old) was 9,125 in March 2019, up from 8,278
the year before, a rate of increase greater than population growth.

It is known that urban food environments, within which people make their
daily food choices, are essential in influencing eating behaviours and
health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to food banks. Although
previous work has explored differences in \emph{accessing} food banks,
such as how some households utilize food banks over short periods of
time while others regularly utilize food banks as longer-term resource
(e.g., Enns et al., 2020), we are not aware of any research that has
focused on the geographic component of accessibility.

The study of place-based geographic accessibility is concerned with
capturing the potential to reach destinations of value using the
transportation network (Páez et al., 2012). Indeed, the Government of
Canada's recent Food Policy has made ``access'' to healthy food a
priority for Canadian communities\footnote{https://www.agr.gc.ca/eng/about-our-department/key-departmental-initiatives/food-policy/the-food-policy-for-canada/?id=1597863791042}
and previous survey research has suggested that such accessibility plays
a key role in user satisfaction with food bank service delivery (Holmes
et al., 2018). However, as with research into the prevalence of food
deserts, accessibility to food banks is unlikely to be evenly
distributed, and variation throughout a city can be expected due to
transportation network characteristics, and the spatial distribution of
food bank locations and the population they are meant to serve.
Furthermore, policy responses to the COVID-19 pandemic likely have added
to the distress of vulnerable households. Non-pharmaceutical
interventions during the pandemic involving restrictions in mobility
have increased the friction of travel, in particular by transit on which
low income populations are more reliant (e.g., DeWeese et al., 2020). At
the same time, the pandemic has created additional stress for the
operators of food banks through disruptions in the supply chain (e.g.,
McKay et al., 2021) as well as concerns surrounding the delivery of
service in safe conditions and possible cancellation of food service
programs.

For this study, we aim to look at how the landscape of food banks and
related services (e.g.~low-cost or free meal service providers)
available in Hamilton, Ontario, changed during the pandemic. Did the
number of open food bank services diminish? If so, what was the
accessibility to food banks before the pandemic from the perspective of
low income households, and how it changed during the pandemic? And
finally, who are most likely to have been impacted by changes in the
accessibility landscape? This paper first looks at the distribution of
food banks and related services before and during the pandemic. Then, we
use the balanced floating catchment area approach of Paez et al. (2019)
to investigate the accessibility situation. For this, we use a fully
disaggregated approach based on parcel-level data. Socio-economic and
demographic data are drawn from the latest Census of Canada (2016),
whereas travel information is from the most recent regional travel
survey from 2016. This paper follows reproducible research
recommendations (see Brunsdon and Comber, 2020), and the research was
conducted using open source tools for transportation analysis (Lovelace,
2021). The code and data necessary to reproduce the analysis are
available in a public repository\footnote{add repository}.

\hypertarget{food-insecurity-and-food-bank-use-in-canada}{%
\section{Food Insecurity and Food Bank Use in
Canada}\label{food-insecurity-and-food-bank-use-in-canada}}

@@

Food banks - sometimes also referred to as `food pantries' and `food
shelves' - originated as a community response to aid those with
inadequate food by voluntarily offering them meals and ingredients
(Loopstra and Tarasuk, 2012; Riches, 2002). The scope and objectives of
food banks can vary by region and by country, and these organizations
can include not only prepared meals and aliments, but also shared spaces
to connect in community gardens and community kitchens (Wakefield et
al., 2013). Although in their origin food banks were meant to be provide
a temporary solution to accommodate those in hunger due to job
retrenchments and economic downfalls since the 1980s, over time they
have evolved into a community practice to secure emergency food supplies
for those in need (Loopstra and Tarasuk, 2012; Wakefield et al., 2013).

In Canada, the number of food banks has steadily increased in the past
few decades (Wakefield et al., 2013). The largest database of food banks
and their use comes from the non-profit association Food Banks Canada
(FBC), which conducts an annual assessment through its affiliated
members. FBC's 2018 Hunger Count report\footnote{https://foodbankscanada.ca/getmedia/241fb659-05f5-44a2-9cef-56f5f51db523/HungerCount-2018\_FINAL\_EN.pdf.aspx?ext=.pdf}
(the most recent available) listed 1,830 member food banks across the
country, and found that Canadians visited food banks 1.1 million times
in March of 2018. Of those accessing food banks, certain population
characteristics tend to be over-represented compared to national totals
from the 2016 Canadian Census of Population. According to FBC's 2018
data, single-adult households represent 45\% of those utilizing food
banks despite making up 28\% of Canada's population, 19\% are
single-parent households (compared to 10\% nationally), and 35\% of
those accessing food bank services are children aged 0-18 even though
their share of Canada's national population is approximately 20\%. In
addition, 59\% of households accessing food banks list social or
disability assistance as their primary source of income. Similarly,
using data from the 2011-2012 CCHS, Tarasuk et al. (2019) found higher
odds of food insecurity amongst households relying on social assistance,
those without a university degree or with children under the age of 18,
and individuals that lived alone, renters, and those identifying as
Aboriginal. While surveys revealed that only 20 to 30 percent of those
experiencing food insecurity were found to frequent food banks in Canada
(Tarasuk et al., 2014), research from Ottawa (Enns et al., 2020) and
Vancouver (Black and Seto, 2020) suggests that long-term users tend to
be older, have health or mobility challenges, live in large households,
and are less likely to have employment income.

Economic stress caused by the COVID-19 pandemic, rising unemployment
rates, and changes in poverty levels have disrupted the food environment
and led to higher rates of food insecurity (Niles et al., 2020).
Pandemic-related food security studies in the US have found a
substantial increase in households experiencing food insecurity for the
first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects
associated with food insecurity, increases in food insecurity rates due
to the pandemic, signal a change in the food environment with potential
damages to population health during the course of the pandemic and
beyond (Niles et al., 2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from FBC showed that 52\% of member food banks reported an increase in
usage in March of 2020 when initial lockdown restrictions were put in
place across much of the country. The pandemic also created significant
staffing issues with 42\% of food banks reporting a reduction in
volunteers. However, 53\% of food banks later reported a decrease in use
into the summer of 2020 which FBC members attributed to emergency
financial support programs from the federal government. Nevertheless,
some of these benefit programs were temporary, which suggests that many
households may again turn to food bank services to meet their needs.

In terms of geography, previous research conducted at the provincial
scale using data from the 2011-2012 CCHS found that the prevalence of
food insecurity ranged across the country from 11.8\% of households in
Ontario to 41\% of households in Nunavut (Tarasuk et al., 2019).
However, although previous research has examined the characteristics of
individuals and households accessing food banks, the locational or
transportation accessibility aspect of food bank access is not well
understood. A wealth of literature examining the food desert concept
suggests that, in addition to socio-economic and demographic factors,
location and transportation networks play a key role in a household's
accessibility to healthy foods (Paez et al., 2010; Vanderlee and L'Abbé,
2017; Widener, 2018). For food banks specifically, previous qualitative
research by Smith et al. (2017) in Ontario has noted that participants
access food banks using a variety of modes of transportation and that
some food bank users have expressed difficulties with transportation for
trips to food bank locations, with one respondent stating ``I wish it
{[}the food bank{]} was a little more centrally located. Because if I
didn't have a bike I'd have to walk it all the way out there and back. I
wonder about people who don't.'' From this, the authors conclude that
``transportation can be challenging, particularly if the food bank is
situated in a remote location.'' To offer greater insight into the role
of transportation and location in food bank accessibility, this research
examines how geographic accessibility to food banks and food services
changed in Hamilton during the COVID-19 pandemic.

@@

\hypertarget{destinations-food-banks-and-food-service-locations}{%
\subsubsection{Destinations: Food Banks and Food Service
Locations}\label{destinations-food-banks-and-food-service-locations}}

The locations of food banks and related food services were obtained from
the Hamilton Public Library's Food Access Guide\footnote{http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf}.
The guide was updated in April of 2021 to indicate any change affected
on the services due to the pandemic. This includes modified business
hours, a need to make reservations before frequenting, and locations
that have completely shut down in consequence. While some food bank
services have a specific a target population, such as prioritizing
family with young children aged between 0 and 3 or accepting only those
providing proof of low-income status through housing and utility
statements, all the food bank services indicated below are designed to
accommodate those in need of food at zero to low cost. With our focus on
food banks and food services that offer free or low-cost meals at
particular locations, we first removed services such as Meals on Wheels
and other food access services such as food box, community kitchens,
student nutrition programs, and shopping and transportation. In
addition, two free meal services held on different days at the same
location were collapsed into a single service point for the
accessibility analysis.

@@

Travel time tables for three modes (car, transit, walking) were computed
using the parcels as the origins and the locations of the food banks as
the destinations. For routing, the package \texttt{r5r} (Pereira et al.,
2021b) was used with a network extract for the City of Hamilton from
OpenStreetMaps and the General Transit Feed Specification from Hamilton
Street Railway, the local transit operator. For routing purposes we used
maximum values of 180 min and 10,000 m walking distance: any destination
that exceeded these thresholds was ignored. The departure time used for
routing was \textbf{TIME}.

@@

Once we obtained travel time tables with population (number of
households) and proportion of trips by mode, we calculated the expected
travel time \(ett\) from each parcel \(i\) to a food bank or food
service location \(j\) as follows: \[
ett_{ij} = p^c_i\cdot tt^c_{ij} + p^t_i\cdot tt^t_{ij} + p^w_i\cdot tt^w_{ij}
\]

\noindent where \(p^k_i\) is the proportion of trips by mode \(k\) in
the TAZ of parcel \(i\), and \(tt^k_{ij}\) is the travel time from
parcel \(i\) to the food bank. In other words, the expected travel time
is the weighted sum of travel times to the food bank, with the weights
given by the expected modal split in the TAZ.

@@

Figure \ref{fig:foodbanks} shows the location of food banks in the City
of Hamilton and their status. Before the pandemic there were 58 of which
14 (24.14\%) closed during the pandemic. As shown in the figure, food
banks tend to be predominantly located in the central parts of the city.
This is not surprising: population density is high there, and it is also
the part of the city where lower income households are more numerous in
absolute and relative terms (see Figure
\ref{fig:low-income-households}). Alas, this is also the part of the
city where most of the closures during the pandemic happened.

@@

To implement the accessibility calculations, we must select a
distance-decay function. In this task we find limited support in the
literature, which is mostly silent on the travel patterns of people who
visit food banks. For this reason, we opt for a simple cumulative
opportunities function as follows: \[
w_{ij}=w_{ji}=
\begin{cases}
1 & \text{ if } ett_{ij}\le \delta\\
0 & \text{ otherwise}
\end{cases}
\] \noindent where \(ett_{ij}\) is the multimodal expected travel time
as described previously, and \(\delta\) is a travel threshold. When the
expected travel time exceeds this threshold, a facility is no longer
considered accessible. Further, the weights are standardized for the
balanced floating catchment area approach.

Figure \ref{fig:sensitivity-analysis} shows the results of conducting
sensitivity analysis of the system-wide accessibility as we vary the
threshold (considering the situation before the pandemic). There is a
clear pattern whereby more strict values of \(\delta\) are associated
with higher levels of system-wide accessibility: while increases in
accessibility that result from decreases in the travel time window might
seem counter-intuitive, this is a result of lower \emph{congestion},
since fewer households are serviced and are thus competing for the same
resources. System-wide accessibility declines with higher values of
\(\delta\): as more households are serviced, congestion grows and the
level of service declines, although this happens at a declining rate. We
are not aware of any standards that explain how long people are expected
to travel for food banks, but we note that in developing countries,
access to drinking water is defined as a source that takes less than 30
minutes to reach (round trip, see UNICEF-WHO, 2019). There is no reason
why people in affluent countries should be expected to travel more for a
basic need such as food. Accordingly, we adopt a threshold of 15 minutes
for the analysis (representing a one-way trip).

@@

Using this threshold, we find that the system-wide accessibility
(interpreted as a provider-to-population ratio) was 0.065 (food banks
and food service locations per low income household in the city) before
COVID-19, but declined to 0.037 during the pandemic. It is striking that
although almost 76\% of food banks remained in operation during the
pandemic, there was a loss of accessibility greater than 43\%,
suggesting the location of food banks and food services plays an
important role in serving those in need.

Turning to the individual food bank and food service locations, the
levels of service offered before and during the pandemic are shown in
Figure \ref{fig:levels-of-service}. The level of service is functionally
the inverse (because each food bank has a ``capacity'' of 1) of the
number of low-income households in the travel-mode weighted travel time
catchment area of the food banks. Higher values mean that a facility is
expected to service fewer households. Conversely, lower values indicate
grater congestion.

The general pattern of the levels of service is similar before and
during the pandemic, with lower values in the center of the city. Three
more peripheral facilities towards the south of the city also have lower
levels of service, presumably because they are meant to service a
relatively large suburban/exurban population. During the pandemic,
however, the levels of service dropped, in some cases quite
substantially. The pattern of the losses in level of service, moreover,
is not uniform. The upper pane of Figure
\ref{fig:levels-of-service-changes} shows that the three peripheral
facilities in the southern suburban/exurban part of the city had low
levels of service to begin with, but did not see major declines during
the pandemic. Turning to the inset map, levels of service deteriorated
more in the central part of the city. However, the loss of level of
service was not as large in the core (where most of the food banks are),
but more marked in the inner ring around the core where facilities may
have faced greater demand from both central city and suburban
populations after the closure of food banks during the pandemic.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{Accessibility-Foodbanks-Hamilton_files/figure-latex/plot-local-i-1} \caption{\label{fig:plot-local-i}Areas with significantly large changes in accessibility from pre-COVID-19 to during COVID-19 according to Local Moran's I.}\label{fig:plot-local-i}
\end{figure}

Discuss the results of aggregating the accessibility by age group (see
Table \ref{tab:accessibility-by-age}). Follow with the analysis of hours
of travel?
